---
title: 'Control Flow'
author: "James M. Flegal"
output: beamer_presentation
---

## Agenda

- Control flow (or alternatively, flow of control)
- if(), for(), and while()
- Avoiding iteration
- Introduction to strings and string operations

## Control flow

- *Control flow* is the order in which individual statements, instructions or function calls of an imperative program are executed or evaluated

- A *control flow statement* is a statement whose execution results in a choice being made as to which of two or more paths should be followed

## Conditionals

- Have the computer decide what to do next
    - Mathematically
\[
|x| = \left\{ \begin{array}{cl} x & \mathrm{if}~x\geq 0 \\
-x &\mathrm{if}~ x < 0\end{array}\right. ~,~ 
\psi(x) = \left\{ \begin{array}{cl} x^2 & \mathrm{if}~|x|\leq 1\\
2|x|-1 &\mathrm{if}~ |x| > 1\end{array}\right.
\]  
    - Computationally
```
if the country code is not "US", 
multiply prices by current exchange rate
```

## if()

- Simplest conditional
```
if (x >= 0) {
  x
} else {
  -x
}
```
- Condition in `if` needs to give _one_ `TRUE` or `FALSE` value
- `else` clause is optional
- one-line actions don't need braces
```
if (x >= 0) x else -x
```

## if()

- `if` can *nested* arbitrarily deep
```
if (x^2 < 1) {
  x^2
} else {
  if (x >= 0) {
    2*x-1
  } else {
     -2*x-1
  }
}
```
- Can get ugly!

## Combining Booleans

- `&` work `|` like `+` or `*` in that they combine terms element-wise
- Flow control wants *one* Boolean value, and to skip calculating what's not needed
- `&&` and `||` give _one_ Boolean, lazily
```{r}
(0 > 0) && (all.equal(42%%6, 169%%13))
```
- This *never* evaluates the complex expression on the right
- Use `&&` and `||` for control, `&` and `|` for subsetting

## Iteration

- Repeat similar actions multiple times

\tiny
```{r}
table.of.logarithms <- vector(length=7,mode="numeric")
table.of.logarithms
for (i in 1:length(table.of.logarithms)) {
  table.of.logarithms[i] <- log(i)
}
table.of.logarithms
```

## for()

```
for (i in 1:length(table.of.logarithms)) {
  table.of.logarithms[i] <- log(i)
}
```

- `for` increments a **counter** (here `i`) along a vector (here `1:length(table.of.logarithms)`) and **loops through** the **body** until it runs through the vector
- That is, it **iterates over** the vector
- Note, there is a better way to do this job!
- Can contain just about anything, including
    - if() clauses
    - other for() loops (nested iteration)

## Nested iteration example

```
c <- matrix(0, nrow=nrow(a), ncol=ncol(b))
if (ncol(a) == nrow(b)) {
  for (i in 1:nrow(c)) {
    for (j in 1:ncol(c)) {
      for (k in 1:ncol(a)) {
        c[i,j] <- c[i,j] + a[i,k]*b[k,j]
      }
    }
  }
} else {
  stop("matrices a and b non-conformable")
}
```

## while()

```
while (max(x) - 1 > 1e-06) {
  x <- sqrt(x)
}
```

- Condition in the argument to `while` must be a single Boolean value (like `if`)
- Body is looped over until the condition is `FALSE` (can loop forever)
- Loop never begins unless the condition starts `TRUE`

## for() vs. while()

- `for()` is better when the number of times to repeat (values to iterate over) is clear in advance
- `while()` is better when you can recognize when to stop once you're there, even if you can't guess it to begin with
- Every `for()` could be replaced with a `while()`
    - Show this on your own!
    
## Avoiding iteration

- `R` has many ways of _avoiding_ iteration, by acting on whole objects
    - Conceptually clearer
    - Leads to simpler code
    - Faster (sometimes a little, sometimes drastically)

## Vectorized arithmetic

- Many languages add 2 vectors using
```
c <- vector(length(a))
for (i in 1:length(a)) {  c[i] <- a[i] + b[i]  }
```
- `R` adds 2 vectors using
```
a+b
```
- Triple `for()` loop for matrix multiplication vs. `a %*% b`

## Advantages of vectorizing

- Clarity: syntax is about _what_ we're doing
- Concision: write less
- Abstraction: syntax hides _how the computer does it_
- Generality: same syntax works for numbers, vectors, arrays, ...
- Speed: modifying big vectors over and over is slow in R; work gets done by optimized low-level code

## Vectorized calculations

- Many functions are set up to vectorize automatically
```{r}
abs(-3:3)
log(1:7)
```

- See also `apply()`

## Vectorized conditions with `ifelse()`

```
ifelse(x^2 > 1, 2*abs(x)-1, x^2)
```
- 1st argument is a Boolean vector, then pick from the 2nd or 3rd vector arguments as `TRUE` or `FALSE`

## What Is Truth?

- 0 counts as `FALSE`; other numeric values count as `TRUE`; the strings "TRUE" and "FALSE" count as you'd hope; most everything else gives an error
- Don't play games here; try to make sure control expressions are getting Boolean values
- Conversely, in arithmetic, `FALSE` is 0 and `TRUE` is 1

\tiny
```{r}
library(datasets)
states <- data.frame(state.x77, abb=state.abb, region=state.region, division=state.division)
mean(states$Murder > 7)
```

## `switch()`

- Simplify nested `if` with `switch()`; give a variable to select on, then a value for each option
```
switch(type.of.summary,
       mean=mean(states$Murder),
       median=median(states$Murder),
       histogram=hist(states$Murder),
       "I don't understand")
```
- Exercise on your own: Set `type.of.summary` to, successively, "mean", "median", "histogram", and "mode", and explain what happens

## Unconditional iteration

\tiny
```
repeat {
  print("Help! I am Dr. Morris Culpepper, trapped in an endless loop!")
}
```

## "Manual" control over iteration

\tiny
```
repeat {
  if (watched) { next() }
  print("Help! I am Dr. Morris Culpepper, trapped in an endless loop!")
  if (rescued) { break() }
}
```

\normalsize

- `break()` exits the loop; `next()` skips the rest of the body and goes back into the loop
- Both work with `for()` and `while()` as well
- Exercise: how would you replace `while()` with `repeat()`?

## Strings and string operations

- Most data we deal with is in character form!
    - web pages can be scraped
    - email can be analyzed for network properties
    - survey responses must be processed and compared
- Even if you only care about numbers, it helps to be able to extract them from text and manipulate them easily.

## Characters vs. Strings

- ***Character*** is a symbol in a written language, specifically what you can enter at a keyboard; letters, numerals, punctuation, space, newlines, etc.
```
'L', 'i', 'n', 'c', 'o', 'l'
```
- ***String*** is a sequence of characters bound together
```
Lincoln
```
- `R` does not have a separate type for characters and strings

\tiny
```{r}
mode("L")
mode("Lincoln")
class("Lincoln")
```

## Making Strings

- Use single or double quotes to construct a string; use `nchar()` to get the length of a single string. Why do we prefer double quotes?

\tiny 
```{r}
"Lincoln"
"Abraham Lincoln"
"Abraham Lincoln's Hat"
"As Lincoln never said, \"Four score and seven beers ago\""
```

## Whitespace

- The space, `" "` is a character; so are multiple spaces `"   "` and the empty string, `""`.
- Some characters are special, so we have "escape characters" to specify them in strings.
    - quotes within strings: `\"`
    - tab: `\t`
    - new line: `\n` (use this when possible)
    - carriage return `\r`

## Character data type

- One of the atomic data types, like `numeric` or `logical`
- Can go into scalars, vectors, arrays, lists, or be the type of a column in a data frame.

\tiny
```{r}
length("Abraham Lincoln's beard")
length(c("Abraham", "Lincoln's", "beard"))
nchar("Abraham Lincoln's beard")
nchar(c("Abraham", "Lincoln's", "beard"))
```

## Character-valued variables

- They work just like others, e.g., with vectors

\tiny
```{r}
president <- "Lincoln"
nchar(president)  # NOT 9
presidents <- c("Fillmore","Pierce","Buchanan","Davis","Johnson")
presidents[3]
presidents[-(1:3)]
```

## Displaying characters

- Know `print()`, of course; `cat()` writes the string directly to the console. If you're debugging, `message()` is preferred syntax in `R`.

\tiny
```{r}
print("Abraham Lincoln")
cat("Abraham Lincoln")
cat(presidents)
message(presidents)
```

## Substring operations

- ***Substring*** is a smaller string from the big string, but still a string in its own right. 
- A string is not a vector or a list, so we ***cannot*** use subscripts like `[[ ]]` or `[ ]` to extract substrings; we use `substr()` instead.

\tiny
```{r}
phrase <- "Christmas Bonus"
substr (phrase, start=8, stop=12)
```

\normalsize
- Can also use `substr` to replace elements

\tiny
```{r}
substr(phrase, 13, 13) <- "g"
phrase
```

## substr() for string vectors

- `substr()` vectorizes over all its arguments

\tiny
```{r}
presidents
substr(presidents,1,2)   # First two characters
substr(presidents,nchar(presidents)-1,nchar(presidents))   # Last two
substr(presidents,20,21)    # No such substrings so return the null string
substr(presidents,7,7)      # Explain!
```

## Dividing strings into vectors

- `strsplit()` divides a string according to key characters, by splitting each element of the character vector `x` at appearances of the pattern `split`.

\tiny
```{r}
scarborough.fair <- "parsley, sage, rosemary, thyme"
strsplit (scarborough.fair, ",")
strsplit (scarborough.fair, ", ")
```

\normalsize
- Pattern is recycled over elements of the input vector

\tiny
```{r}
strsplit (c(scarborough.fair, "Garfunkel, Oates", "Clement, McKenzie"), ", ")
```

## Combining vectors into strings

- Converting one variable type to another is called ***casting***

\tiny
```{r}
as.character(7.2)            # Obvious
as.character(7.2e12)         # Obvious
as.character(c(7.2,7.2e12))  # Obvious
as.character(7.2e5)          # Not quite so obvious
```

## Building strings from multiple parts

- The `paste()` function is very flexible!
- With one vector argument, works like `as.character()`

```{r}
paste(41:45)
```

## Building strings from multiple parts

- With 2 or more vector arguments, combines them with recycling

\tiny
```{r}
paste(presidents,41:45)
paste(presidents,c("R","D"))  # Not historically accurate!
paste(presidents,"(",c("R","D"),41:45,")")
```

## Building strings from multiple parts

- Changing the separator between pasted-together terms

\tiny
```{r}
paste(presidents, " (", 41:45, ")", sep="_")
paste(presidents, " (", 41:45, ")", sep="")
```

\normalsize
- What happens if you give `sep` a vector?

## More complicated example of recycling

- Exercise: Convince yourself of why this works as it does

\tiny
```{r}
paste(c("HW","Lab"),rep(1:11,times=rep(2,11)))
```

## Condensing multiple strings

- Producing one big string

\tiny
```{r}
paste(presidents, " (", 41:45, ")", sep="", collapse="; ")
```

\normalsize
- Default value of `collapse` is `NULL` -- that is, it won't use it

## Function for writing regression formulas

- `R` has a standard syntax for models: outcome and predictors.

\tiny
```{r}
my.formula <- function(dep,indeps,df) {
  rhs <- paste(colnames(df)[indeps], collapse="+")
  return(paste(colnames(df)[dep], " ~  ", rhs, collapse=""))
}
my.formula(2,c(3,5,7),df=state.x77)
```

## General search

- Use `grep()` to find which strings have a matching search term
- Reconstituting, make one long string, then split the words
- Counting words with `table()`
- Need to learn how to work with text patterns and not just constants
- Searching for text patterns using regular expressions

## Summary

- `if`, nested `if`, `switch`
- Iteration with `for` and `while`
- Avoiding iteration with whole-object ("vectorized") operations
- Text is data, just like everything else
