---
title: 'More Data Structures'
author: "James M. Flegal"
output: beamer_presentation
---

## Agenda
- Arrays
- Matrices
- Lists
- Dataframes
- Structures of structures

## Arrays

- Many data structures in R are made by adding bells and whistles to vectors, so "vector structures"
- Most useful is **arrays**
```{r}
x <- c(7, 8, 10, 45)
x.arr <- array(x,dim=c(2,2))
x.arr
```

- `dim` says how many rows and columns; filled by columns
- Can have $3, 4, \ldots n$ dimensional arrays; `dim` is a length-$n$ vector

## Arrays

- Some properties of the array
```{r}
dim(x.arr)
is.vector(x.arr)
is.array(x.arr)
```

## Arrays

```{r}
typeof(x.arr)
str(x.arr)
attributes(x.arr)
```
- `typeof()` returns the type of the _elements_
- `str()` gives the **structure**: here, a numeric array, with two dimensions, both indexed 1--2, and then the actual numbers
- Exercise: try all these with `x`

## Accessing and operating on arrays

- Can access a 2-D array either by pairs of indices or by the underlying vector
```{r}
x.arr[1,2]
x.arr[3]
```

## Accessing and operating on arrays

- Omitting an index means "all of it"
```{r}
x.arr[c(1:2),2]
x.arr[,2]
```

## Functions on arrays

- Using a vector-style function on a vector structure will go down to the underlying vector, _unless_ the function is set up to handle arrays specially
```{r}
which(x.arr > 9)
```

## Accessing and operating on arrays

- Many functions _do_ preserve array structure
```{r}
y <- -x
y.arr <- array(y,dim=c(2,2))
y.arr + x.arr
```

- Others specifically act on each row or column of the array separately
```{r}
rowSums(x.arr)
```

- See more of this idea later

## Example: Price of houses in PA

- Census data for California and Pennsylvania on housing prices, by Census "tract"

\tiny
```{r}
calif_penn <- read.csv("http://www.stat.cmu.edu/~cshalizi/uADA/13/hw/01/calif_penn_2011.csv")
penn <- calif_penn[calif_penn[,"STATEFP"]==42,]
coefficients(lm(Median_house_value ~ Median_household_income, data=penn))
```
\normalsize

- Fit a simple linear model, predicting median house price from median household income

## Example: Price of houses in PA

- Census tracts 24--425 are Allegheny county
- Tract 24 has a median income of \$14,719; actual median house value is \$34,100 --- is that above or below what's predicted?
```{r}
34100 < -26206.564 + 3.651*14719
```

- Tract 25 has income \$48,102 and house price \$155,900
```{r}
155900 < -26206.564 + 3.651*48102
```

- What about tract 26?

## Example: Price of houses in PA

- _Could_ just keep plugging in numbers like this, but that's
    - boring and repetitive
    - error-prone (what if I forget to change the median income, or drop a minus sign from the intercept?)
    - obscure if we come back to our work later (what _are_ these numbers?)

## Use variables and names

\tiny
```{r}
penn.coefs <- coefficients(lm(Median_house_value ~ Median_household_income, data=penn))
penn.coefs
```

```{r}
allegheny.rows <- 24:425
allegheny.medinc <- penn[allegheny.rows,"Median_household_income"]
allegheny.values <- penn[allegheny.rows,"Median_house_value"]
allegheny.fitted <- penn.coefs["(Intercept)"]+penn.coefs["Median_household_income"]*allegheny.medinc
```

## Use variables and names

\tiny
```{r}
plot(x=allegheny.fitted, y=allegheny.values, xlab="Model-predicted median house values",
     ylab="Actual median house values", xlim=c(0,5e5),ylim=c(0,5e5))
abline(a=0,b=1,col="grey")
```

## Resource allocation example

- Factory makes cars and trucks, using labor and steel
    - a car takes 40 hours of labor and 1 ton of steel
    - a truck takes 60 hours and 3 tons of steel
    - resources: 1600 hours of labor and 70 tons of steel each week

## Matrices

- Matrix is a specialization of a 2D array

```{r}
factory <- matrix(c(40,1,60,3),nrow=2)
is.array(factory)
is.matrix(factory)
```

- Could also specify `ncol`, and/or `byrow=TRUE` to fill by rows.
- Element-wise operations with the usual arithmetic and comparison operators (e.g., `factory/3`)
- Compare whole matrices with `identical()` or `all.equal()`

## Matrix multiplication

- Gets a special operator

```{r}
six.sevens <- matrix(rep(7,6),ncol=3)
six.sevens
factory %*% six.sevens # [2x2] * [2x3]
```

- What happens if you try `six.sevens %*% factory`?


## Multiplying matrices and vectors

- Numeric vectors can act like proper vectors
```{r}
output <- c(10,20)
factory %*% output
output %*% factory
```

- R silently casts the vector as either a row or a column matrix


## Matrix operators

- Transpose
```{r}
t(factory)
```

- Determinant
```{r}
det(factory)
```

## The diagonal

- The `diag()` function can extract the diagonal entries of a matrix
```{r}
diag(factory)
```

- It can also _change_ the diagonal
```{r}
diag(factory) <- c(35,4)
factory
```

- Re-set it for later
```{r}
diag(factory) <- c(40,3)
```

## Creating a diagonal or identity matrix

```{r}
diag(c(3,4))
diag(2)
```

## Inverting a matrix

```{r}
solve(factory)
factory %*% solve(factory)
```

## Why's it called "solve" anyway?

- Solving the linear system $\mathbf{A}\vec{x} = \vec{b}$ for $\vec{x}$
```{r}
available <- c(1600,70)
solve(factory,available)
factory %*% solve(factory,available)
```

## Names in matrices

- Can name either rows or columns or both, with `rownames()` and `colnames()`
- Just character vectors, and we use the same function to get and to set their values
- Names help us understand what we're working with
- Can be used to coordinate different objects

## Names in matrices

```{r}
rownames(factory) <- c("labor","steel")
colnames(factory) <- c("cars","trucks")
factory
available <- c(1600,70)
names(available) <- c("labor","steel")
```

## Names in matrices

\tiny
```{r}
output <- c(20,10)
names(output) <- c("trucks","cars")
factory %*% output # But we've got cars and trucks mixed up!
factory %*% output[colnames(factory)]
all(factory %*% output[colnames(factory)] <= available[rownames(factory)])
```

\normalsize
- Last lines don't have to change if we add motorcycles as output or rubber and glass as inputs (abstraction again)


## Doing the same thing to each row or column

- Take the mean: `rowMeans()`, `colMeans()`: input is matrix, output is vector.  Also `rowSums()`, etc.
- `summary()`: vector-style summary of column

\tiny
```{r}
colMeans(factory)
summary(factory)
```

## `apply()`

- Takes 3 arguments: the array or matrix, then 1 for rows and 2 for columns, then name of the function to apply to each

```{r}
rowMeans(factory)
apply(factory,1,mean)
```

- What would `apply(factory,1,sd)` do?
- More on `apply()` later

## Lists

- Sequence of values, _not_ necessarily all of the same type

```{r}
my.distribution <- list("exponential",7,FALSE)
my.distribution
```

- Most of what you can do with vectors you can also do with lists

## Accessing pieces of lists

- Can use `[ ]` as with vectors
- Or use `[[ ]]`, but only with a single index
    - `[[ ]]` drops names and structures, `[ ]` does not
```{r}
is.character(my.distribution)
is.character(my.distribution[[1]])
my.distribution[[2]]^2
```

- What happens if you try `my.distribution[2]^2`?
- What happens if you try `[[ ]]` on a vector?

## Expanding and contracting lists

- Add to lists with `c()` (also works with vectors)
```{r}
my.distribution <- c(my.distribution,7)
my.distribution
```

## Expanding and contracting lists

- Chop off the end of a list by setting the length to something smaller (also works with vectors)

\tiny
```{r}
length(my.distribution)
length(my.distribution) <- 3
my.distribution
```

## Naming list elements

- We can name some or all of the elements of a list

\tiny
```{r}
names(my.distribution) <- c("family","mean","is.symmetric")
my.distribution
my.distribution[["family"]]
my.distribution["family"]
```

## Naming list elements

- Lists have a special short-cut way of using names, `$` (which removes names and structures)
\tiny
```{r}
my.distribution[["family"]]
my.distribution$family
```

## Naming list elements

1. Creating a list with names
2. Adding named elements
3. Removing a named list element, by assigning it the value `NULL`

\tiny
```{r}
another.distribution <- list(family="gaussian",mean=7,sd=1,is.symmetric=TRUE)

my.distribution$was.estimated <- FALSE
my.distribution[["last.updated"]] <- "2011-08-30"

my.distribution$was.estimated <- NULL
```


## Key-Value pairs

- Lists give us a way to store and look up data by _name_, rather than by _position_
- Really useful programming concept with many names: **key-value pairs**, **dictionaries**, **associative arrays**, **hashes**
- If all our distributions have components named `family`, we can look that up by name, without caring where it is in the list

## Dataframes

- Dataframe = the classic data table, $n$ rows for cases, $p$ columns for variables
- Lots of the really-statistical parts of R presume data frames
    - `penn` from last time was really a dataframe
- Not just a matrix because *columns can have different types*
- Many matrix functions also work for dataframes (`rowSums()`, `summary()`, `apply()`)
    - but no matrix multiplying dataframes, even if all columns are numeric

## Dataframes

\tiny 
```{r}
a.matrix <- matrix(c(35,8,10,4),nrow=2)
colnames(a.matrix) <- c("v1","v2")
a.matrix
a.matrix[,"v1"]  # Try a.matrix$v1 and see what happens
```

## Dataframes

\tiny 
```{r}
a.data.frame <- data.frame(a.matrix,logicals=c(TRUE,FALSE))
a.data.frame
a.data.frame$v1
a.data.frame[,"v1"]
a.data.frame[1,]
colMeans(a.data.frame)
```

## Adding rows and columns

- Can add rows or columns to an array or data-frame with `rbind()` and `cbind()`, but be careful about forced type conversions

\tiny
```{r}
rbind(a.data.frame,list(v1=-3,v2=-5,logicals=TRUE))
rbind(a.data.frame,c(3,4,6))
```

## Structures of Structures

- So far, every list element has been a single data value
- List elements can be other data structures, e.g., vectors and matrices

\tiny
```{r}
plan <- list(factory=factory, available=available, output=output)
plan$output
```

\normalsize
- Internally, a dataframe is basically a list of vectors

## Structures of Structures

- List elements can even be other lists
    - which may contain other data structures
        - including other lists ... which may contain other data structures

- This **recursion** lets us build arbitrarily complicated data structures from the basic ones
- Most complicated objects are (usually) lists of data structures

## Example: Eigenstuff

- `eigen()` finds eigenvalues and eigenvectors of a matrix
- Returns a list of a vector (the eigenvalues) and a matrix (the eigenvectors)

\tiny
```{r}
eigen(factory)
class(eigen(factory))
```

## Example: Eigenstuff

- With complicated objects, you can access parts of parts (of parts...)

\tiny
```{r}
factory %*% eigen(factory)$vectors[,2]
eigen(factory)$values[2] * eigen(factory)$vectors[,2]
eigen(factory)$values[2]
eigen(factory)[[1]][[2]] # NOT [[1,2]]
```

## Creating an example dataframe

\tiny
```{r}
library(datasets)
states <- data.frame(state.x77, abb=state.abb, region=state.region, division=state.division)
```

\normalsize
- `data.frame()` is combining here a pre-existing matrix (`state.x77`), a vector of characters (`state.abb`), and two vectors of qualitative categorical variables (**factors**; `state.region`, `state.division`)
- Column names are preserved or guessed if not explicitly set

## Creating an example dataframe

\tiny
```{r}
colnames(states)
states[1,]
```

## Dataframe access

- By row and column index
```{r}
states[49,3]
```
- By row and column names
```{r}
states["Wisconsin","Illiteracy"]
```

## Dataframe access

- All of a row:

\tiny
```{r}
states["Wisconsin",]
```

## Dataframe access

- All of a column:
```{r}
head(states[,3])
head(states[,"Illiteracy"])
head(states$Illiteracy)
```

## Dataframe access

- Rows matching a condition:
```{r}
states[states$division=="New England", "Illiteracy"]
states[states$region=="South", "Illiteracy"]
```

## Dataframe access

- Parts or all of the dataframe can be assigned to

\tiny
```{r}
summary(states$HS.Grad)
states$HS.Grad <- states$HS.Grad/100
summary(states$HS.Grad)
states$HS.Grad <- 100*states$HS.Grad
```


## with()

- What percentage of literate adults graduated HS?

\tiny
```{r}
head(100*(states$HS.Grad/(100-states$Illiteracy)))
```

\normalsize
- `with()` takes a data frame and evaluates an expression "inside" it

\tiny
```{r}
with(states, head(100*(HS.Grad/(100-Illiteracy))))
```

## Data arguments

- Lots of functions take `data` arguments, and look variables up in that data frame
```{r}
plot(Illiteracy~Frost, data=states)
```

<small>$R^2 =0.45$, $p \approx {10}^{-7}$</small>

## Summary
- Arrays add multi-dimensional structure to vectors
- Matrices act like you'd hope they would
- Lists let us combine different types of data
- Dataframes are hybrids of matrices and lists, for classic tabular data
- Recursion lets us build complicated data structures out of the simpler ones
